---
title: "NHANES Application"
author: "Cole Manschot"
date: "2025-07-15"
output:
  html_document:
    toc: true           # Enable table of contents
    toc_float: true     # Make TOC float (sticky sidebar)
    toc_depth: 3        # Depth of headers included in TOC
    theme: paper       # Optional: Bootstrap theme for nicer style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warnings = FALSE)
library(dplyr)
library(tidyr)
library(ggplot2)
library(mice)
```

## Models 

We will consider five models for the data application. 
These are motivated by connecting dietary measures to outcomes that relate to clinically relevant outcomes and deficiencies. 

The models considered are as follows: 

| Outcome | Covariate | Clinical Relevance |
|---|---|---|
| Vitamin D | Calcium Intake | Vitamin D facilitates calcium absorption |
| Resting Heart Rate | Caffeine Intake | Caffeine can affect the heart rate and cardiovascular function |
| HDL Cholesterol | Saturated Fat Intake | HDL cholesterol is associated cardiovascular disease and saturated fat is influenced HDL levels |
| Insulin | Alcohol Consumption | Insulin resistance is associated with numerous co-morbidities and alcohol may impact insulin sensitivity | 
| Folate | Folate Food | Low levels of iron contribute to anemia risk and is related to iron intake | 

These five models relate relevant health outcomes to dietary factors which individuals may have some level of control over. 
However, accurate measurements of dietary intake are often confounded by recall bias and measurement error from mapping foods and proportions to nutrient intake. 

## NHANES Data

These models allow us to use three data sets from the NHANES (and one additional data set for demographic information). 
This simplifies the data application and makes it more easily reproducible. 

```{r}
analysis_data = read.csv("~/Documents/ETS_PCA/NHANES-Analysis/analysis_data_with_errors.csv")
```

## Visualize Error-Prone (Observed) vs. Error-Free (Simulated)

```{r}
plot_data = analysis_data |> 
  select(SEQN, X1:X5) |> 
  pivot_longer(cols = X1:X5, names_to = "Model", values_to = "ValX") |> 
  mutate(Model = sub(pattern = "X", replacement = "", x = Model)) |>
  left_join(
    analysis_data |> 
      select(SEQN, XSTAR1:XSTAR5) |> 
      pivot_longer(cols = XSTAR1:XSTAR5, 
                   names_to = "Model", 
                   values_to = "UnvalX") |> 
      mutate(Model = sub(pattern = "XSTAR", 
                         replacement = "", 
                         x = Model))
  )

plot_data |> 
  ggplot(aes(x = UnvalX, y = ValX)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  theme_minimal() + 
  facet_wrap(~Model, scales = "free")
```

### Simulate different validation study designs 

```{r}
## Convert factor covariates, subset to necessary columns
analysis_data = analysis_data |> 
  mutate(RIDRETH1 = as.factor(RIDRETH1), 
         DMDEDUC2 = as.factor(DMDEDUC2)) |> 
  select(Y1:X5, XSTAR1:XSTAR5, RIAGENDR, RIDAGEYR, RIDRETH1, DMDEDUC2)  
Z = c("RIAGENDR", "RIDAGEYR") # , "RIDRETH1", "DMDEDUC2")

## Initialize dataframe to hold estimates from the 5 models
fits = data.frame()

## Set validation study size
n = 250
```

```{r srs}
## For reproducibility 
set.seed(918)

## Simple random sampling 
V_srs = auditDesignR::sample_srs(phI = nrow(analysis_data), 
                                 phII = n)
des_srs = analysis_data |> 
  bind_cols(data.frame(V = V_srs)) |> 
  mutate(X1 = ifelse(test = V == 1, yes = X1, no = NA), 
         X2 = ifelse(test = V == 1, yes = X2, no = NA), 
         X3 = ifelse(test = V == 1, yes = X3, no = NA), 
         X4 = ifelse(test = V == 1, yes = X4, no = NA), 
         X5 = ifelse(test = V == 1, yes = X5, no = NA))

## Loop over j = 1, ..., 5 to impute and fit each model
for (j in 1:5) {
  ### Imputation model depends on the validation study design and number of imputations
  ### Which variables go into the imputation model 
  imp_mod_vars = c(paste0("X", j), paste0("XSTAR", j), Z) #### All include Xj, Xj*, Z
  imp_mod_vars = c(imp_mod_vars, paste0("Y", j)) #### Multiple imputation adds Yj
  
  ### Impute and fit model 
  #### Multiple imputation
  mice_dat = mice(m = 75,
                  data = des_srs[, imp_mod_vars], 
                  method = "norm",
                  printFlag = FALSE)
  
  #### Fit analysis model to the imputed data (separately) 
  after_imp_fit = with(data = mice_dat, 
                       expr = glm(formula = as.formula(paste0("Y", j, "~", "X", j, "+", paste(Z, collapse = "+"))), 
                                  family = "gaussian"))
  
  #### Pool the analysis models from each imputation
  pool_imp_fit = summary(pool(after_imp_fit))
  
  #### Save coefficient estimates
  fits = fits |> 
    bind_rows(data.frame(cbind(model = j, design = "SRS", pool_imp_fit)))
}
```

```{r ETS(X1*)}
## ETS on X1*
justXSTAR1 = data.frame(row_num = 1:nrow(analysis_data), 
                        XSTAR1 = analysis_data$XSTAR1) 
    
## Order ascendingly by X*1
justXSTAR1 = justXSTAR1[order(justXSTAR1$XSTAR1, decreasing = FALSE), ]
### Only validate smallest n/2 X*1
smallest_XSTAR1 = justXSTAR1$row_num[1:(250 / 2)]

## Re-order descendingly by first principal component
justXSTAR1 = justXSTAR1[order(justXSTAR1$XSTAR1, decreasing = TRUE), ]
### Only validate smallest n/2 residuals
largest_XSTAR1 = justXSTAR1$row_num[1:(250 / 2)]

## Define validation indicator
V_etsXSTAR1 = c(smallest_XSTAR1, largest_XSTAR1)
des_etsXSTAR1 = analysis_data |> 
  bind_cols(data.frame(V = 1:nrow(analysis_data) %in% V_etsXSTAR1)) |> 
  mutate(X1 = ifelse(test = V == 1, yes = X1, no = NA), 
         X2 = ifelse(test = V == 1, yes = X2, no = NA), 
         X3 = ifelse(test = V == 1, yes = X3, no = NA), 
         X4 = ifelse(test = V == 1, yes = X4, no = NA), 
         X5 = ifelse(test = V == 1, yes = X5, no = NA))

## Loop over j = 1, ..., 5 to impute and fit each model
for (j in 1:5) {
  ### Imputation model depends on the validation study design and number of imputations
  ### Which variables go into the imputation model 
  imp_mod_vars = c(paste0("X", j), paste0("XSTAR", j), Z) #### All include Xj, Xj*, Z
  imp_mod_vars = unique(c(imp_mod_vars, "XSTAR1")) #### ETS-X1 adds X1* 
  imp_mod_vars = c(imp_mod_vars, paste0("Y", j)) #### Multiple imputation adds Yj
  
  ### Impute and fit model 
  #### Multiple imputation
  mice_dat = mice(m = 75,
                  data = des_etsXSTAR1[, imp_mod_vars], 
                  method = "norm",
                  printFlag = FALSE)
  
  #### Fit analysis model to the imputed data (separately) 
  after_imp_fit = with(data = mice_dat, 
                       expr = glm(formula = as.formula(paste0("Y", j, "~", "X", j, "+", paste(Z, collapse = "+"))), 
                                  family = "gaussian"))
  
  #### Pool the analysis models from each imputation
  pool_imp_fit = summary(pool(after_imp_fit))
  
  #### Save coefficient estimates
  fits = fits |> 
    bind_rows(data.frame(cbind(model = j, design = "ETS (X1*)", pool_imp_fit)))
}
```

```{r ETS(PC1*)}
## ETS on PC1*
pc = analysis_data |> 
  dplyr::select(XSTAR1:XSTAR5) |> 
  princomp(cor = TRUE) ### use correlation, since Xs are on different scales
pc1 = data.frame(row_num = 1:nrow(analysis_data), 
                 pc = pc$scores[, 1]) ## extract the first principal component

## Order ascendingly by first principal component
pc1 = pc1[order(pc1$pc, decreasing = FALSE), ]
### Only validate smallest n/2 principal components
smallest_pc1 = pc1$row_num[1:(250 / 2)]

## Re-order descendingly by first principal component
pc1 = pc1[order(pc1$pc, decreasing = TRUE), ]
### Only validate smallest n/2 residuals
largest_pc1 = pc1$row_num[1:(250 / 2)]

## Define validation indicator
V_etsPCstar1 = c(smallest_pc1, largest_pc1)
des_etsPCstar1 = analysis_data |> 
  bind_cols(data.frame(V = 1:nrow(analysis_data) %in% V_etsPCstar1)) |> 
  mutate(X1 = ifelse(test = V == 1, yes = X1, no = NA), 
         X2 = ifelse(test = V == 1, yes = X2, no = NA), 
         X3 = ifelse(test = V == 1, yes = X3, no = NA), 
         X4 = ifelse(test = V == 1, yes = X4, no = NA), 
         X5 = ifelse(test = V == 1, yes = X5, no = NA))

## Re-order by row number to merge into data
des_etsPCstar1$pc1 = as.numeric(pc1[order(pc1$row_num, decreasing = FALSE), 2])

## Loop over j = 1, ..., 5 to impute and fit each model
for (j in 1:5) {
  ### Imputation model depends on the validation study design and number of imputations
  ### Which variables go into the imputation model 
  imp_mod_vars = c(paste0("X", j), paste0("XSTAR", j), Z) #### All include Xj, Xj*, Z
  imp_mod_vars = c(imp_mod_vars, "pc1") #### ETS-PCA adds pc 
  imp_mod_vars = c(imp_mod_vars, paste0("Y", j)) #### Multiple imputation adds Yj
  
  ### Impute and fit model 
  #### Multiple imputation
  mice_dat = mice(m = 75,
                  data = des_etsPCstar1[, imp_mod_vars], 
                  method = "norm",
                  printFlag = FALSE)
  
  #### Fit analysis model to the imputed data (separately) 
  after_imp_fit = with(data = mice_dat, 
                       expr = glm(formula = as.formula(paste0("Y", j, "~", "X", j, "+", paste(Z, collapse = "+"))), 
                                  family = "gaussian"))
  
  #### Pool the analysis models from each imputation
  pool_imp_fit = summary(pool(after_imp_fit))
  
  #### Save coefficient estimates
  fits = fits |> 
    bind_rows(data.frame(cbind(model = j, design = "ETS (PC1*)", pool_imp_fit)))
}
```

```{r}
# Function to parse LaTex labels 
parse.labels <- function(x) parse(text = x)
library(latex2exp)
fits |> 
  dplyr::filter(grepl(pattern = "X[1-5]", x = term)) |> 
  mutate(Model = factor(x = model, 
                        levels = c(1, 2, 3, 4, 5), 
                        labels = c(TeX("$Y_1 \\sim X_1 + \\bf{Z}$"), 
                                   TeX("$Y_2 \\sim X_2 + \\bf{Z}$"),
                                   TeX("$Y_3 \\sim X_3 + \\bf{Z}$"), 
                                   TeX("$Y_4 \\sim X_4 + \\bf{Z}$"),
                                   TeX("$Y_5 \\sim X_5 + \\bf{Z}$"))),
         Design = factor(x = design, 
                         levels = c("SRS", "ETS (X1*)", "ETS (PC1*)"), 
                         labels = c("SRS", TeX("ETS-$X_1^*$"), TeX("ETS-$PC_1^*$"))))  |> 
  ggplot(aes(x = Design, y = estimate, color = Design)) + 
  geom_point() + 
  geom_errorbar(aes(ymin = estimate - 1.96 * std.error, ymax = estimate + 1.96 * std.error)) + 
  theme_minimal(base_size = 14) + 
  ggthemes::scale_color_colorblind(guide = "none") + 
  facet_wrap(~Model,
             scales = "free", 
             labeller = labeller(Model = label_parsed)) + 
  xlab(TeX("Validation Study Design", bold = TRUE)) + 
  ylab(TeX("Coefficient Estimate on $X_j$", bold = TRUE)) + 
  scale_x_discrete(labels = parse.labels) + 
  theme(strip.background = element_rect(fill = "black"), 
        strip.text = element_text(color = "white"), 
        legend.title = element_text(face = "bold"), 
        legend.position = "top")
```
```{r, echo = FALSE}
## Save it 
if(grepl(pattern = "lotspes", x = getwd())) {
 ggsave(filename = "~/Documents/ETS_PCA/Plots/NHANES_Analysis.pdf", 
       device = "pdf", 
       width = 10, 
       height = 5) 
}
```

```{r}
fits |> 
  dplyr::filter(grepl(pattern = "X[1-5]", x = term)) |> 
  mutate(Model = factor(x = model, 
                        levels = c(1, 2, 3, 4, 5), 
                        labels = c(TeX("$Y_1 \\sim X_1 + \\bf{Z}$"), 
                                   TeX("$Y_2 \\sim X_2 + \\bf{Z}$"),
                                   TeX("$Y_3 \\sim X_3 + \\bf{Z}$"), 
                                   TeX("$Y_4 \\sim X_4 + \\bf{Z}$"),
                                   TeX("$Y_5 \\sim X_5 + \\bf{Z}$"))),
         Design = factor(x = design, 
                         levels = c("SRS", "ETS (X1*)", "ETS (PC1*)"), 
                         labels = c("SRS", TeX("ETS-$X_1^*$"), TeX("ETS-$PC_1^*$"))))  |> 
  dplyr::mutate(lb = estimate - 1.96 * std.error, 
                ub = estimate + 1.96 * std.error, 
                ci_width = ub - lb) |> 
  dplyr::select(Model, Design, ci_width) |> 
  ggplot(aes(x = Model, y = ci_width, fill = Design)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 1)) + 
  theme_minimal(base_size = 14) + 
  ggthemes::scale_fill_colorblind(labels = parse.labels) + 
  scale_x_discrete(labels = parse.labels) + 
  xlab(TeX("Model", bold = TRUE)) + 
  ylab(TeX("Width of the 95% Confidence Interval for $\\beta_j$", bold = TRUE)) + 
  theme(strip.background = element_rect(fill = "black"), 
        strip.text = element_text(color = "white"), 
        legend.title = element_text(face = "bold"), 
        legend.position = "top")
```

```{r}
fits = fits |> 
  dplyr::filter(grepl(pattern = "X[1-5]", x = term)) |> 
  mutate(Model = factor(x = model, 
                        levels = c(1, 2, 3, 4, 5), 
                        labels = c(TeX("$Y_1 \\sim X_1 + \\bf{Z}$"), 
                                   TeX("$Y_2 \\sim X_2 + \\bf{Z}$"),
                                   TeX("$Y_3 \\sim X_3 + \\bf{Z}$"), 
                                   TeX("$Y_4 \\sim X_4 + \\bf{Z}$"),
                                   TeX("$Y_5 \\sim X_5 + \\bf{Z}$"))),
         Design = factor(x = design, 
                         levels = c("SRS", "ETS (X1*)", "ETS (PC1*)"), 
                         labels = c("SRS", TeX("ETS-$X_1^*$"), TeX("ETS-$PC_1^*$"))))  |> 
  dplyr::mutate(lb = estimate - 1.96 * std.error, 
                ub = estimate + 1.96 * std.error, 
                ci_width = ub - lb)
fits |> 
  dplyr::left_join(
    fits |> 
      dplyr::filter(design == "SRS") |> 
      dplyr::select(model, ci_width) |> 
      dplyr::rename(srs_ci_width = ci_width)
  ) |> 
  dplyr::mutate(width_ratio = ci_width / srs_ci_width) |> 
  ggplot(aes(x = Model, y = width_ratio, fill = Design)) + 
  geom_bar(stat = "identity", position = position_dodge(width = 1)) + 
  theme_minimal(base_size = 14) + 
  ggthemes::scale_fill_colorblind(labels = parse.labels) + 
  scale_x_discrete(labels = parse.labels) + 
  xlab(TeX("Model", bold = TRUE)) + 
  ylab(TeX("Width of the 95% Confidence Interval for $\\beta_j$", bold = TRUE)) + 
  theme(strip.background = element_rect(fill = "black"), 
        strip.text = element_text(color = "white"), 
        legend.title = element_text(face = "bold"), 
        legend.position = "top")
```

```{r}
## Create the plot 
fits |> 
  dplyr::filter(grepl(pattern = "X[1-5]", x = term)) |> 
  mutate(Model = factor(x = model, 
                        levels = c(1, 2, 3, 4, 5), 
                        labels = c(TeX("$Y_1 \\sim X_1 + \\bf{Z}$"), 
                                   TeX("$Y_2 \\sim X_2 + \\bf{Z}$"),
                                   TeX("$Y_3 \\sim X_3 + \\bf{Z}$"), 
                                   TeX("$Y_4 \\sim X_4 + \\bf{Z}$"),
                                   TeX("$Y_5 \\sim X_5 + \\bf{Z}$"))),
         Design = factor(x = design, 
                         levels = c("SRS", "ETS (X1*)", "ETS (PC1*)"), 
                         labels = c("SRS", TeX("ETS-$X_1^*$"), TeX("ETS-$PC_1^*$"))))  |> 
  dplyr::mutate(lb = estimate - 1.96 * std.error, 
                ub = estimate + 1.96 * std.error, 
                ci_width = ub - lb) |> 
  dplyr::select(Model, Design, ci_width) |> 
  dplyr::group_by(Model) |> 
  dplyr::mutate(minWidth = min(ci_width), 
                maxWidth = max(ci_width)) |> 
  ggplot(aes(x = Model, 
             y = ci_width, 
             color = Design)) + 
  geom_segment(aes(x = Model, y = minWidth,
                   xend = Model, yend = maxWidth),
               color = "#aeb6bf",
               linewidth = 4.5, 
               alpha = 0.5) +
  geom_point(size = 4) + 
  facet_grid(cols = vars(Model),
             scales = "free") +
  theme_minimal(base_size = 14) + 
  ggthemes::scale_color_colorblind(labels = parse.labels) + 
  xlab(TeX("Model of $Y_j \\sim X_j$", bold = TRUE)) + 
  ylab(TeX("Empirical Efficiency of Coefficient Estimate on $X_j$ (Log-Transformed)", bold = TRUE)) + 
  scale_x_discrete(labels = parse.labels) + 
  theme(strip.background = element_rect(fill = "black"), 
        strip.text = element_text(color = "white"), 
        legend.title = element_text(face = "bold"), 
        legend.position = "top", 
        panel.spacing = unit(1, "lines")) + 
  coord_flip() 
```